Transient Variability In Vapor Intrusion And The Factors That Influence It

### 2018-12-13 tex/main.tex
Added some new figures

% additional packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{array,multirow,graphicx}
\usepackage{comment}
\graphicspath{
  {./../figures/},
  {./../figures/air-exchange-rate/},
  {./../figures/kde-iacc-pressure/},
  {./../figures/preferential-pathway-sensitivity/},
  {./../figures/transient-response/}
}
\usepackage{lineno}
\linenumbers
% macros

@@

\begin{figure}[htb!]
  \caption{Sensitivity analysis of IACC dependence on indoor/outdoor pressure difference for cases featuring a preferential pathway (\ref{fig:ss-sensitivity-analysis-pp}) and without a preferential pathway (\ref{fig:ss-sensitivity-analysis-no-pp}). Results compared with field data from ASU house.}
  \label{fig:ss-sensitivity-analysis}
  % "Overview" plot
  \begin{subfigure}{0.55\textwidth}
    \caption{preferential pathway present. Sensitivity to the presence of a gravel sub-base and contamination in the preferential pathway considered.}
    \label{fig:ss-sensitivity-analysis-pp}
    \includegraphics[width=\textwidth]{yes-pp.pdf}
  \end{subfigure}
  % Violinplot figure
  \begin{subfigure}{0.55\textwidth}
    \caption{No preferential pathway present. Sensitivitiy to the presence of a gravel sub-base considered.}
    \label{fig:ss-sensitivity-analysis-no-pp}
    \includegraphics[width=\textwidth]{no-pp.pdf}
  \end{subfigure}
\end{figure}

@@

\begin{comment}

Clearly, PPs pose a major problem for site investigations, in particular because they may be difficult or even impossible to anticipate or uncover; even at a well studied site like the ASU house it took years to discover the PP.
Therefore, there is a need to consider other indications that a PP is present.

% Sensitivity analysis plot showing when a PP is relevant
\begin{figure}[htb!]
  \caption{Attenuation factor relative to contaminant vapor in equilibrium with groundwater as a function of indoor-outdoor pressure difference. The effects of a preferential pathway, that is either contaminated or uncontaminated as well as that of having a gravel sub-base vs. uniform sandy clay soil are considered.}
  \label{fig:model_results_sensitivity_analysis}
  % Upper left
  \begin{subfigure}[t]{0.45\textwidth}
    \caption{ }
    \label{fig:model_results_sensitivity_analysis_0}
    \includegraphics[width=\textwidth]{ss_pp_sensitivity_analysis_0.png}
  \end{subfigure}
  % Upper right
  \begin{subfigure}[t]{0.45\textwidth}
    \caption{ }
    \label{fig:model_results_sensitivity_analysis_1}
    \includegraphics[width=\textwidth]{ss_pp_sensitivity_analysis_1.png}
  \end{subfigure}
  % Middle left
  \begin{subfigure}[t]{0.45\textwidth}
    \caption{ }
    \label{fig:model_results_sensitivity_analysis_2}
    \includegraphics[width=\textwidth]{ss_pp_sensitivity_analysis_2.png}
  \end{subfigure}
  % Middle right
  \begin{subfigure}[t]{0.45\textwidth}
    \caption{ }
    \label{fig:model_results_sensitivity_analysis_3}
    \includegraphics[width=\textwidth]{ss_pp_sensitivity_analysis_3.png}
  \end{subfigure}
  % Lower left
  \begin{subfigure}[t]{0.45\textwidth}
    \caption{ }
    \label{fig:model_results_sensitivity_analysis_4}
    \includegraphics[width=\textwidth]{ss_pp_sensitivity_analysis_4.png}
  \end{subfigure}
  % Lower right
  \begin{subfigure}[t]{0.45\textwidth}
    \caption{ }
    \label{fig:model_results_sensitivity_analysis_5}
    \includegraphics[width=\textwidth]{ss_pp_sensitivity_analysis_5.png}
  \end{subfigure}
\end{figure}

One way has been suggested by the situation at the ASU house, where there clearly was a PP containing contaminant vapor, and the PP exited into a gravel (permeable) sub-base.
This situation can be examined using the VI model introduced earlier by examining how IACC is impacted by various factors over a range of indoor-outdoor pressure difference values.
Instead of representing IACC as absolute concentration, we now non-dimensionalize with respect to the vapor in equilibrium with the groundwater contaminant concentration.
This non-dimensionalized property is commonly called the attenuation factor and is denoted by $\alpha_\mathrm{gw}$.
The result of a sensitivity analysis relative to several factors is seen in Figure \ref{fig:model_results_sensitivity_analysis}.
These examine scenarios in which the indoor-outdoor pressure difference is a driver for VI.

In this figure, all the graphs in the left column show results for a scenario, in which the house has a gravel sub-base, with the remaining surrounding soil assumed to be sandy clay.
All the right column graphs features only sandy clay soil, directly in contact with the slab.
Figure \ref{fig:model_results_sensitivity_analysis_0} considers an uncontaminated PP ($\chi = 0$), i.e. the PP delivers clean air to beneath the slab.
In \ref{fig:model_results_sensitivity_analysis_1} there is no PP present at all, giving a reference state for a "normal" VI scenario.
\ref{fig:model_results_sensitivity_analysis_2} features a contaminated PP ($\chi = 1$), i.e. the PP delivers contaminant vapor at a concentration characteristic of underlying groundwater.

In terms of conditions that may lead to significant temporal transients in $\alpha_\mathrm{gw}$, driven by indoor-outdoor pressure difference, most of the cases in this analysis show negative results.
There are only two combinations of factors where significant changes in $\alpha_\mathrm{gw}$ accompany the changes in indoor-outdoor pressure difference, both involving the permeable (gravel) sub-base into which a PP enters.
In \ref{fig:model_results_sensitivity_analysis_2}, the PP is filled with contaminant vapor whereas figure \ref{fig:model_results_sensitivity_analysis} involves the PP only delivering "clean" air to the gravel sub-base.
When the building is overpressurized (positive indoor-outdoor pressure difference), the two cases exhibit similar behavior, i.e. $\alpha_\mathrm{gw}$ is low as expected.

When the building is underpressurized, the uncontaminated PP case involves barely any increase in $\alpha_\mathrm{gw}$ while there is a very significant increase in $\alpha_\mathrm{gw}$ in the contaminated PP case.
For the contaminated PP scenario, this is quite easy to understand, as the extra air flow provided by the PP into the gravel sub-base very effectively disperses contaminant vapor across the sub-base and the contaminant entry rate is increased.

When the PP is provides clean air, the increase in air flow into the sub-base with depressurization is the same.
However, the resulting increase in air entry rate does not bring in more contaminant as clean air is being drawn in.
This means that most of the contaminant entry rate is still only controlled by the usual process of diffusion through the soil, giving a result that is actually similar to that for the "no PP and with gravel sub-base" scenario.
Thus, for a PP to be a significant contributor to contaminant entry at a VI site the conditions such as featured in the bottom left case are required, i.e. a permeable sub-base and the PP must be a source of contaminant.

Various techniques for determining the potential for sewer gas entry into a house are described by \citeauthor{nielsen_remediation_2017-1}\cite{nielsen_remediation_2017-1}.
Collecting contaminant vapor samples from nearby manholes is also recommended, however one should be aware that contaminant vapors can potentially travel very long distances in a sewer system\cite{mchugh_evidence_2017,roghani_occurrence_2018,riis_vapor_2010}.

% Models specific to ASU & North Island
\begin{figure}[htb!]
  \caption{Attenuation factor vs. indoor-outdoor pressure difference when considering VI scenarios similar to the ASU house and the North Island NAS site (\ref{fig:ss_asu} and \ref{fig:ss_north_island} respectively).}
  \label{fig:model_results_asu_north_island}
  % ASU
  \begin{subfigure}[t]{0.45\textwidth}
    \caption{ }
    \label{fig:ss_asu}
    \includegraphics[width=\textwidth]{ss_asu.png}
  \end{subfigure}
  % North Island
  \begin{subfigure}[t]{0.45\textwidth}
    \caption{ }
    \label{fig:ss_north_island}
    \includegraphics[width=\textwidth]{ss_north_island.png}
  \end{subfigure}
\end{figure}

At the ASU house, after the PP was closed, there was still roughly an order of magnitude variation in IACC, the reasons for which Figure \ref{fig:model_results_sensitivity_analysis} does not address.
Likewise, there was more than two orders of magnitude observed at the North Island site over the indoor-outdoor pressure difference ranges simulated despite there not being a PP present.
Therefore, it is of interest to examine these sites a bit further using the model.

The foundation crack at the ASU house seems by all accounts to be quite small, roughly 180x1 cm and located close to the PP\cite{guo_vapor_2015}.
This crack is smaller than the one assumed in the model used in Figure \ref{fig:model_results_sensitivity_analysis}.
The influence of indoor-outdoor pressure difference is explored for a case in which a smaller crack area is assumed.
We also consider the effects of having a significantly lower air exchange rate, 0.1 vs. the regular 0.5 per hour.

The results of the case shown in Figure \ref{fig:model_results_sensitivity_analysis_2} is shown in Figure \ref{fig:ss_asu}.
Even though there is an increase in $\alpha_\mathrm{gw}$ with increased underpressurization, it is no where near the roughly one order of magnitude recorded at the ASU house after the PP was closed.
So while crack dimensions can clearly influence $\alpha_\mathrm{gw}$, they cannot be used to infer a bigger role for pressure difference in temporal variations in $\alpha_\mathrm{gw}$ or IACC.
The influence of air exchange rate on $\alpha_\mathrm{gw}$ predictions is also shown in Figure \ref{fig:model_results_asu_north_island}.
Clearly, the value of air exchange rate at steady-state conditions will only influence absolute concentrations in the structure.
Ultimately, the reason for the transient variability in IACC during the post-CPM period at the ASU house remains elusive, as a Pearson's r analysis between IACC and various factors do not yield any explanation.

In the case of the North Island site, variations in IACC are clearly not related to the existence of a PP.
The soil underlying the North Island site is more sand-like\cite{hosangadi_high-frequency_2017}.
The significance of this is that sand soil is much more permeable than sandy clay soil, which means that $\alpha_\mathrm{gw}$ will be much more sensitive to changes in indoor-outdoor pressure difference.
Thus, another "normal" VI scenario was run but this time with sand soil instead of sandy clay, the results of which may be seen in Figure \ref{fig:ss_north_island}.
Under these conditions, $\alpha_\mathrm{gw}$ changes significantly with pressurization, spanning roughly two orders of magnitude.
This is consistent with what was observed at North Island, as there the span of IACC values was roughly the same across this range of indoor-outdoor pressure difference values.
This simulation, together with the Pearson $r = -0.64$ for North Island in Figure \ref{fig:kde-asu-nas} suggest that under conditions where the soil around a structure is relatively permeable, indoor-outdoor pressure difference can be a significant driving forces for transient changes in IACC.
Therefore, recording the indoor-outdoor pressure difference at a site (in particular if the soil is permeable) may offer insight into the potential for transient changes in IACC (some good methods for doing this are suggested by \citeauthor{nielsen_remediation_2017-1}\cite{nielsen_remediation_2017-1}).

% Transient part
\subsubsection{Transient Modeling}

Analyzing the impact of a PP using steady-state simulations such as those in Figures \ref{fig:model_results_sensitivity_analysis} and does not revealing about how quickly IACC can change in the presence of a PP.
A transient simulation of a VI scenario that is characterized by a PP has been performed, where only the indoor-outdoor pressure difference was temporally changed.
The pressure difference is changed in a sinusoidal fashion, defined by \eqref{eq:p_transient},
\begin{equation}
  \label{eq:p_transient}
  p_\mathrm{in/out} = 10 \sin{(2 \pi t)} - 0.5
\end{equation}
where $t$ is given in days and $p_\mathrm{in/out}$ is the indoor-outdoor pressure difference in Pa.
Figure \ref{fig:transient_simulation} shows the results of this simulation.

\begin{figure}
  \caption{Transient response of TCE in indoor air (as attenuation factor) and in the surrounding soil \& gravel sub-base for a PP VI scenario subject to sinusoidal indoor-outdoor pressure fluctuations.
  \ref{fig:alpha_response} shows the attenuation factor and indoor-outdoor pressure across time.
  Figures \ref{fig:soil_30_hr} \& \ref{fig:soil_42_hr} show the contaminant concentration (normalized to the contaminant vapor in equilibrium with groundwater) in the surrounding soil \& gravel sub-base at different depths and times.}
  \label{fig:transient_simulation}
  % Transient response figure
  \begin{subfigure}[t]{0.75\textwidth}
    \caption{ }
    \label{fig:alpha_response}
    \includegraphics[width=\textwidth]{alpha_transient.png}
  \end{subfigure}
  % Soil t = 30 hr ...
  \begin{subfigure}[c]{0.45\textwidth}
    \caption{t = 30 hr}
    \label{fig:soil_30_hr}
    \includegraphics[width=\textwidth]{tri_layer_soil_concs1.png}
  \end{subfigure}
  % Soil t = 42 hr ...
  \begin{subfigure}[c]{0.45\textwidth}
    \caption{t = 42 hr}
    \label{fig:soil_42_hr}
    \includegraphics[width=\textwidth]{tri_layer_soil_concs2.png}
  \end{subfigure}
  % Color map
  \begin{subfigure}[c]{0.05\textwidth}
    \phantomcaption{ }
    %\label{fig:cmap}
    \includegraphics[width=\textwidth]{cmap.png}
  \end{subfigure}
\end{figure}

Figure \ref{fig:alpha_response} shows the IACC response, given as attenuation factor relative to contaminant vapor in equilibrium with groundwater, and the indoor-outdoor pressure difference with time.
What is apparent is that $\alpha_\mathrm{gw}$ closely follows the indoor-outdoor pressure difference, with the max./min. $\alpha_\mathrm{gw}$ being reached roughly 0.5 hours after the indoor-outdoor pressure difference reaches its max./min..
The reader should be aware that the assumption that the indoor air space can be modeled as a CST overestimates how dynamic this process is as a real house probably would not be perfectly mixed on as a short time scale as this.
Regardless, this suggests that VI sites characterized by PPs are highly dynamic and IACC's may change quickly in response to changing indoor-outdoor pressure differences, consistent with what was observed at the ASU house\cite{holton_temporal_2013}.
Again, the dynamics of the indoor-outdoor pressure difference at a VI site can offer insight into the potential for transient behavior of IACC.

Figures \ref{fig:soil_30_hr} and \ref{fig:soil_42_hr} show the contaminant concentration (normalized to the contaminant vapor in equilibrium with groundwater) in the soil and gravel sub-base at three different depths at times of 30 and 42 hr respectively corresponding to roughly the maximum in overpressurization and depressurization respectively.
The top layer is 3 m above groundwater and immediately under the house foundation (indicated by the black square).
The middle and bottom layers are 2 and 1 m above groundwater respectively.
The black streamlines show the contaminant transport flow paths.
The region closest to the PP is expanded in the circular insets.

The impact of the PP on the contaminant vapor concentration in the subsurface is apparent in these figures, as one easily sees how (especially) the gravel sub-base is filled up.
The contaminant vapor concentration is the highest closest to the PP (which is located at the edge of the gravel sub-base) and decreases with increased distance from the PP.
Note that a very small amount of contaminant vapor leaves the gravel sub-base, as the sandy clay has a relatively large resistance to transport compared with gravel.
This makes it difficult to detect the contribution of the PP outside the foundation footprint.
However, this may not always be true, as vapor contaminant from the PP would disperse more easily into more permeable soils.

It should also be noted that the vapor contaminant emanating from the PP in this model is relatively high concentration; the vapor contaminant concentration is equal to that in equilibrium with the groundwater source.
This does not always seem to be the case, as VI sites that have uncovered PPs find that the vapor contaminant emanating from the PP can be lower than that\cite{guo_vapor_2015}.
Therefore, it may be much more difficult than it may appear in Figures \ref{fig:soil_30_hr} and \ref{fig:soil_42_hr} to localize a vapor contaminant hotspot as a way to determining the presence of a PP at a particular VI site.

The insert circles in figures \ref{fig:soil_30_hr} and \ref{fig:soil_42_hr} also show that for the simulated time period, only a small amount of the contaminant enters the structure (as evident by the blue low concentration zone around the perimeter), yet a very large increase in $\alpha_\mathrm{gw}$ occurs.
The concentration in the rest of the sub-base is unchanged.
This suggests that once a PP has deposited contaminant underneath a structure, it may persist for a long time period, as pointed out by \citeauthor{guo_vapor_2015}\cite{guo_vapor_2015}.

\end{comment}

@@

>>>>

% additional packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{array,multirow,graphicx}
\usepackage{comment}
\graphicspath{
  {./../figures/},
  {./../figures/air-exchange-rate/},
  {./../figures/kde-iacc-pressure/},
  {./../figures/preferential-pathway-sensitivity/},
  {./../figures/transient-response/},
  {./../figures/multivariate_analysis/},
  {./../figures/rate_of_change/},
}
\usepackage{lineno}
\linenumbers
% macros

@@

\begin{figure}[htb!]
  \caption{Sensitivity analysis of IACC dependence on indoor/outdoor pressure difference for cases featuring a preferential pathway (\ref{fig:ss-sensitivity-analysis-pp}) and without a preferential pathway (\ref{fig:ss-sensitivity-analysis-no-pp}). Results compared with field data from ASU house.}
  \label{fig:ss-sensitivity-analysis}
  % "Overview" plot
  \begin{subfigure}{0.45\textwidth}
    \caption{preferential pathway present. Sensitivity to the presence of a gravel sub-base and contamination in the preferential pathway considered.}
    \label{fig:ss-sensitivity-analysis-pp}
    \includegraphics[width=\textwidth]{yes-pp.pdf}
  \end{subfigure}
  % Violinplot figure
  \begin{subfigure}{0.45\textwidth}
    \caption{No preferential pathway present. Sensitivitiy to the presence of a gravel sub-base considered.}
    \label{fig:ss-sensitivity-analysis-no-pp}
    \includegraphics[width=\textwidth]{no-pp.pdf}
  \end{subfigure}
\end{figure}

@@

\subsubsection{Multivariate Analysis}

\begin{figure}[htb!]
  \caption{Seasonal distribution of relevant VI parameters and indoor air contaminant concentration at the ASU site. Only period after land drain was closed considered.}
  \label{fig:asu_season}
  \includegraphics[width=\textwidth]{asu_season.pdf}
\end{figure}

\begin{figure}[htb!]
  \caption{Seasonal distribution of relevant VI parameters and indoor air contaminant concentration at the Indianapolis site. Only PCE considered.}
  \label{fig:indianapolis_season}
  \includegraphics[width=\textwidth]{indianapolis_season.pdf}
\end{figure}

Pressure alone does not seem adequately be able to predict the variability observed at the ASU and Indianapolis VI sites, necessitating a multivariate analysis to better understand the distribution of indoor air contaminant concentration.
% Add references motivating the choice of parameters
This is acheived by examining the distribution various parameters recorded at these sites, and how they relate to eachother.
It is known that VI potential differs across the seasons, thus the seasonal distribution of each parameter is considered.

@@

\subsubsection{Rates of Change in Indoor Air Contaminant Concentration}

\begin{figure}[htb!]
  \caption{}
  \label{fig:}
  % "Overview" plot
  \begin{subfigure}{0.45\textwidth}
    \caption{}
    \label{fig:}
    \includegraphics[width=\textwidth]{rate_of_change_season_asu_pp_open.pdf}
  \end{subfigure}
  % Violinplot figure
  \begin{subfigure}{0.45\textwidth}
    \caption{}
    \label{fig:}
    \includegraphics[width=\textwidth]{rate_of_change_season_asu_pp_closed.pdf}
  \end{subfigure}
  % Indianapolis
  \begin{subfigure}{0.45\textwidth}
    \caption{}
    \label{fig:}
    \includegraphics[width=\textwidth]{rate_of_change_season_indianapolis.pdf}
  \end{subfigure}
  % Indianapolis
  \begin{subfigure}{0.45\textwidth}
    \caption{}
    \label{fig:}
    \includegraphics[width=\textwidth]{rate_of_change_simulations.pdf}
  \end{subfigure}
\end{figure}
